### 2019-12-15 ic-journal/multivariate.tex
text

With the formal definitions of \multivariate substitution and
\multivariate (weakly guarded) contexts, now we are ready to formally
define \multivariate CCS equations/contractions and their (unique)
solutions. Consider a system of equation $\{X_i = E_i\}_{i\in I}$
(Def.~\ref{def:equation}), or its expanded form (here we suppose $I = [1,n]
\in \mathbb{N}$, i.e.~a finite list):
\begin{equation*}
  \begin{cases}
    &X_1 = E_1[\til X] \\
    &X_2 = E_2[\til X] \\
    & \cdots \\
    &X_n = E_n[\til X]
  \end{cases}
\end{equation*}
Consider its two essential ingredients:
\begin{itemize}
\item $\til X = (X_1, X_2, \ldots, X_n)$: a list of equation variables;
\item $\til E = (E_1, E_2, \ldots, E_n)$: a list of CCS contexts 
  with possible occurrences of free variables in $\til X$.
\end{itemize}
The two lists $\til X$ and $\til E$ must have the same
length, and  the variables in $\til X$ should  be distinct.
Furthermore,  $\til E$ does not contain
 free variables that are not in $\til X$.
\hlD{Finally, } for each
$E_i$, the set of its bound variables must be \emph{disjoint} from
$\til X$. 
% \hl{(In practice, however, one can always
% rename all recursion binders in the equations to satisfy this
% requirement without changing the semantics of equations.)}
These requirements yield the fomal
definition of \multivariate CCS equation:
\begin{alltt}
   \HOLTokenTurnstile{} \HOLConst{CCS_equation} \HOLFreeVar{Xs} \HOLFreeVar{Es} \HOLSymConst{\HOLTokenEquiv{}}
      \HOLConst{ALL_DISTINCT} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenConj{}} \HOLConst{LENGTH} \HOLFreeVar{Es} \HOLSymConst{\ensuremath{=}} \HOLConst{LENGTH} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenConj{}}
      \HOLConst{EVERY} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{e}. \HOLConst{FV} \HOLBoundVar{e} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{set} \HOLFreeVar{Xs}\ensuremath{)} \HOLFreeVar{Es} \HOLSymConst{\HOLTokenConj{}}
      \HOLConst{EVERY} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{e}. \HOLConst{DISJOINT} \ensuremath{(}\HOLConst{BV} \HOLBoundVar{e}\ensuremath{)} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Xs}\ensuremath{)}\ensuremath{)} \HOLFreeVar{Es}\hfill{[CCS_equation_def]}
\end{alltt}

Now consider (formally) what is a solution $\til P$ of CCS equations. First of
all, the definition should be parametrized on a binary CCS
relation $\R$ such as $\sim$ and $\rcontr$, so that a single definition
can be used to represent solutions of all kinds of CCS
\hlD{equations.} Then,  ${\til P} \ \R\  {\til E}\{\til
P/\til X\}$ must hold:
\begin{equation*}
  \begin{cases}
    &P_1 \ \R\ E_1\{\til P/\til X\} \\
    &P_2 \ \R\ E_2\{\til P/\til X\} \\
    & \cdots \\
    &P_n \ \R\ E_n\{\til P/\til X\}
  \end{cases}
\end{equation*}
Furthermore, each $P_i$ should be a pure process, i.e.~having no free
variable. Finally, the set of bound variables is disjoint with
$\til X$. (This disjointness requirement is optional but makes many proofs
easier.) Putting all together,
below is the formal definition of a solution of \multivariate CCS
equations:\footnote{If $R$ is a binary relation of CCS processes,
  \HOLinline{\HOLConst{LIST_REL}\\\;\HOLFreeVar{R}} is the same binary relation but for lists of
  CCS processes with the same length.
 % Furthermore,  \HOLinline{\HOLConst{LIST_REL}\\\;\HOLFreeVar{R}\\\;\HOLFreeVar{A}\\\;\HOLFreeVar{B}}
  implicitly   implies that the two lists $A$ and $B$ have the same length.
}
\begin{alltt}
   \HOLTokenTurnstile{} \HOLConst{CCS_solution} \HOLFreeVar{R} \HOLFreeVar{Xs} \HOLFreeVar{Es} \HOLFreeVar{Ps} \HOLSymConst{\HOLTokenEquiv{}}
      \HOLConst{ALL_PROC} \HOLFreeVar{Ps} \HOLSymConst{\HOLTokenConj{}} \HOLConst{EVERY} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{e}. \HOLConst{DISJOINT} \ensuremath{(}\HOLConst{BV} \HOLBoundVar{e}\ensuremath{)} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Xs}\ensuremath{)}\ensuremath{)} \HOLFreeVar{Ps} \HOLSymConst{\HOLTokenConj{}}
      \HOLConst{LIST_REL} \HOLFreeVar{R} \HOLFreeVar{Ps} \ensuremath{(}\HOLConst{MAP} \ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLFreeVar{Es}\ensuremath{)}\hfill{[CCS_solution_def]}
\end{alltt}
Note that the two logical terms ``\HOLinline{\HOLConst{CCS_solution}\\\;\HOLFreeVar{R}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Es}\\\;\HOLFreeVar{Ps}}'' and ``\HOLinline{\HOLFreeVar{Ps}\\\;\HOLSymConst{\HOLTokenIn{}}\\\;\HOLConst{CCS_solution}\\\;\HOLFreeVar{R}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Es}}'' in
HOL have the same meaning. The latter form suggests that,
``\HOLinline{\HOLConst{CCS_solution}\\\;\HOLFreeVar{R}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Es}}'' is actually a set containing all
solutions of ``\HOLinline{\HOLConst{CCS_equation}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Es}}''. Then the unique-solution
theorems can be understood thus: any two (syntactically different) elements
in the solution set are bisimilar.

>>>>

With the formal definitions of \multivariate substitution and
\multivariate (weakly guarded) contexts, now we are ready to formally
define \multivariate CCS equations/contractions and their (unique)
solutions. Consider a system of equation $\{X_i = E_i\}_{i\in I}$
(Def.~\ref{def:equation}), or its expanded form (here we suppose $I = [1,n]
\in \mathbb{N}$, i.e.~a finite list):
\begin{equation*}
  \begin{cases}
    &X_1 = E_1[\til X] \\
    &X_2 = E_2[\til X] \\
    & \cdots \\
    &X_n = E_n[\til X]
  \end{cases}
\end{equation*}
Consider its two essential ingredients:
\begin{itemize}
\item $\til X = (X_1, X_2, \ldots, X_n)$: a list of equation variables;
\item $\til E = (E_1, E_2, \ldots, E_n)$: a list of CCS contexts 
  with possible occurrences of free variables in $\til X$.
\end{itemize}
The two lists $\til X$ and $\til E$ should have the same
length, and  the variables in $\til X$ should  be distinct.
Furthermore,  $\til E$ does not contain
 free variables that are not in $\til X$.
\hlD{Finally, } for each
$E_i$, the set of its bound variables should be \emph{disjoint} from
$\til X$. 
% \hl{(In practice, however, one can always
% rename all recursion binders in the equations to satisfy this
% requirement without changing the semantics of equations.)}
These requirements yield the fomal
definition of \multivariate CCS equation:
\begin{alltt}
   \HOLTokenTurnstile{} \HOLConst{CCS_equation} \HOLFreeVar{Xs} \HOLFreeVar{Es} \HOLSymConst{\HOLTokenEquiv{}}
      \HOLConst{ALL_DISTINCT} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenConj{}} \HOLConst{LENGTH} \HOLFreeVar{Es} \HOLSymConst{\ensuremath{=}} \HOLConst{LENGTH} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenConj{}}
      \HOLConst{EVERY} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{e}. \HOLConst{FV} \HOLBoundVar{e} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{set} \HOLFreeVar{Xs}\ensuremath{)} \HOLFreeVar{Es} \HOLSymConst{\HOLTokenConj{}}
      \HOLConst{EVERY} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{e}. \HOLConst{DISJOINT} \ensuremath{(}\HOLConst{BV} \HOLBoundVar{e}\ensuremath{)} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Xs}\ensuremath{)}\ensuremath{)} \HOLFreeVar{Es}\hfill{[CCS_equation_def]}
\end{alltt}

Now consider (formally) what is a solution $\til P$ of CCS equations. First of
all, the definition should be parametrized on a binary CCS
relation $\R$ such as $\sim$ and $\rcontr$, so that a single definition
can be used to represent solutions of all kinds of CCS
\hlD{equations.} Then,  ${\til P} \ \R\  {\til E}\{\til
P/\til X\}$  should hold:
\begin{equation*}
  \begin{cases}
    &P_1 \ \R\ E_1\{\til P/\til X\} \\
    &P_2 \ \R\ E_2\{\til P/\til X\} \\
    & \cdots \\
    &P_n \ \R\ E_n\{\til P/\til X\}
  \end{cases}
\end{equation*}
Furthermore, each $P_i$ should be a pure process, i.e.~having no free
variable. Finally, the set of bound variables is disjoint with
$\til X$. (This disjointness requirement is optional but makes many proofs
easier.) Putting all together,
below is the formal definition of a solution of \multivariate CCS
equations:\footnote{If $R$ is a binary relation of CCS processes,
  \HOLinline{\HOLConst{LIST_REL}\\\;\HOLFreeVar{R}} is the same binary relation but for lists of
  CCS processes with the same length.
 % Furthermore,  \HOLinline{\HOLConst{LIST_REL}\\\;\HOLFreeVar{R}\\\;\HOLFreeVar{A}\\\;\HOLFreeVar{B}}
  implicitly   implies that the two lists $A$ and $B$ have the same length.
}
\begin{alltt}
   \HOLTokenTurnstile{} \HOLConst{CCS_solution} \HOLFreeVar{R} \HOLFreeVar{Xs} \HOLFreeVar{Es} \HOLFreeVar{Ps} \HOLSymConst{\HOLTokenEquiv{}}
      \HOLConst{ALL_PROC} \HOLFreeVar{Ps} \HOLSymConst{\HOLTokenConj{}} \HOLConst{EVERY} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{e}. \HOLConst{DISJOINT} \ensuremath{(}\HOLConst{BV} \HOLBoundVar{e}\ensuremath{)} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Xs}\ensuremath{)}\ensuremath{)} \HOLFreeVar{Ps} \HOLSymConst{\HOLTokenConj{}}
      \HOLConst{LIST_REL} \HOLFreeVar{R} \HOLFreeVar{Ps} \ensuremath{(}\HOLConst{MAP} \ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLFreeVar{Es}\ensuremath{)}\hfill{[CCS_solution_def]}
\end{alltt}
Note that the two logical terms ``\HOLinline{\HOLConst{CCS_solution}\\\;\HOLFreeVar{R}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Es}\\\;\HOLFreeVar{Ps}}'' and ``\HOLinline{\HOLFreeVar{Ps}\\\;\HOLSymConst{\HOLTokenIn{}}\\\;\HOLConst{CCS_solution}\\\;\HOLFreeVar{R}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Es}}'' in
HOL have the same meaning. The latter form suggests that,
``\HOLinline{\HOLConst{CCS_solution}\\\;\HOLFreeVar{R}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Es}}'' is actually a set containing all
solutions of ``\HOLinline{\HOLConst{CCS_equation}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Es}}''. Then the unique-solution
theorems can be understood thus: any two (syntactically different) elements
in the solution set are bisimilar.

### 2019-12-15 ic-journal/sos.tex
text

A useful facility exploiting the interplay
between HOL4 and Standard ML (\hl{which} follows an idea of Nesi~\cite{Nesi:1992ve})
 is a recursive ML function that takes a CCS process and returns a theorem indicating all
  direct transitions of the process. (If the input process is
 infinitely branching, the function will not terminate, however.)
For instance, we know that the process $(a.\nil | \bar{a}.\nil)$ has three
\hl{immediate derivatives given by the following transitions}:
$(a.\nil | \bar{a}.\nil) \overset{a}{\longrightarrow}
(\nil | \bar{a}.\nil)$, $(a.\nil | \bar{a}.\nil)
\overset{\bar{a}}{\longrightarrow} (a.\nil | \nil)$ and $(a.\nil | \bar{a}.\nil)
\overset{\tau}{\longrightarrow} (\nil | \nil)$.
To completely describe all possible transitions of the process, the
following two facts must be proved: (1) there are indeed the three
  transitions mentioned above; (2) there is no other
  transition. For large CCS processes it is surprisingly tedious to manually derive
all the possible transitions and prove the non-existence of other transitions.
This shows the usefulness of appealing to an ML function
\texttt{CCS\_TRANS\_CONV} that is designed to automate the whole process.
For instance, taking the input $(a.\nil | \bar{a}.\nil)$ the function returns the
following theorem which describes all its one-step transitions:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{\HOLTokenInputAct} \HOLStringLit{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{\ensuremath{\mathbf{0}}} \HOLSymConst{\ensuremath{\mid}} \HOLConst{\HOLTokenOutputAct} \HOLStringLit{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{\ensuremath{\mathbf{0}}} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E} \HOLSymConst{\HOLTokenEquiv{}}
   \ensuremath{(}\HOLFreeVar{u} \HOLSymConst{\ensuremath{=}} \HOLConst{\HOLTokenInputAct} \HOLStringLit{a} \HOLSymConst{\HOLTokenConj{}} \HOLFreeVar{E} \HOLSymConst{\ensuremath{=}} \HOLConst{\ensuremath{\mathbf{0}}} \HOLSymConst{\ensuremath{\mid}} \HOLConst{\HOLTokenOutputAct} \HOLStringLit{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{\ensuremath{\mathbf{0}}} \HOLSymConst{\HOLTokenDisj{}} \HOLFreeVar{u} \HOLSymConst{\ensuremath{=}} \HOLConst{\HOLTokenOutputAct} \HOLStringLit{a} \HOLSymConst{\HOLTokenConj{}} \HOLFreeVar{E} \HOLSymConst{\ensuremath{=}} \HOLConst{\HOLTokenInputAct} \HOLStringLit{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{\ensuremath{\mathbf{0}}} \HOLSymConst{\ensuremath{\mid}} \HOLConst{\ensuremath{\mathbf{0}}}\ensuremath{)} \HOLSymConst{\HOLTokenDisj{}}
   \HOLFreeVar{u} \HOLSymConst{\ensuremath{=}} \HOLSymConst{\ensuremath{\tau}} \HOLSymConst{\HOLTokenConj{}} \HOLFreeVar{E} \HOLSymConst{\ensuremath{=}} \HOLConst{\ensuremath{\mathbf{0}}} \HOLSymConst{\ensuremath{\mid}} \HOLConst{\ensuremath{\mathbf{0}}}
\end{alltt}

>>>>

A useful facility exploiting the interplay
between HOL4 and Standard ML (\hl{which} follows an idea of Nesi~\cite{Nesi:1992ve})
 is a recursive ML function that takes a CCS process and returns a theorem indicating all
  direct transitions of the process. (If the input process is
 infinitely branching, the function will not terminate, however.)
For instance, we know that the process $(a.\nil | \bar{a}.\nil)$ has three
\hl{immediate derivatives given by the following transitions}:
$(a.\nil | \bar{a}.\nil) \overset{a}{\longrightarrow}
(\nil | \bar{a}.\nil)$, $(a.\nil | \bar{a}.\nil)
\overset{\bar{a}}{\longrightarrow} (a.\nil | \nil)$ and $(a.\nil | \bar{a}.\nil)
\overset{\tau}{\longrightarrow} (\nil | \nil)$.
To completely describe all possible transitions of the process, the
following two facts have to be proved: (1) there are indeed the three
  transitions mentioned above; (2) there is no other
  transition. For large CCS processes it is surprisingly tedious to manually derive
all the possible transitions and prove the non-existence of other transitions.
This shows the usefulness of appealing to an ML function
\texttt{CCS\_TRANS\_CONV} that is designed to automate the whole process.
For instance, taking the input $(a.\nil | \bar{a}.\nil)$ the function returns the
following theorem which describes all its one-step transitions:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{\HOLTokenInputAct} \HOLStringLit{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{\ensuremath{\mathbf{0}}} \HOLSymConst{\ensuremath{\mid}} \HOLConst{\HOLTokenOutputAct} \HOLStringLit{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{\ensuremath{\mathbf{0}}} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E} \HOLSymConst{\HOLTokenEquiv{}}
   \ensuremath{(}\HOLFreeVar{u} \HOLSymConst{\ensuremath{=}} \HOLConst{\HOLTokenInputAct} \HOLStringLit{a} \HOLSymConst{\HOLTokenConj{}} \HOLFreeVar{E} \HOLSymConst{\ensuremath{=}} \HOLConst{\ensuremath{\mathbf{0}}} \HOLSymConst{\ensuremath{\mid}} \HOLConst{\HOLTokenOutputAct} \HOLStringLit{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{\ensuremath{\mathbf{0}}} \HOLSymConst{\HOLTokenDisj{}} \HOLFreeVar{u} \HOLSymConst{\ensuremath{=}} \HOLConst{\HOLTokenOutputAct} \HOLStringLit{a} \HOLSymConst{\HOLTokenConj{}} \HOLFreeVar{E} \HOLSymConst{\ensuremath{=}} \HOLConst{\HOLTokenInputAct} \HOLStringLit{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{\ensuremath{\mathbf{0}}} \HOLSymConst{\ensuremath{\mid}} \HOLConst{\ensuremath{\mathbf{0}}}\ensuremath{)} \HOLSymConst{\HOLTokenDisj{}}
   \HOLFreeVar{u} \HOLSymConst{\ensuremath{=}} \HOLSymConst{\ensuremath{\tau}} \HOLSymConst{\HOLTokenConj{}} \HOLFreeVar{E} \HOLSymConst{\ensuremath{=}} \HOLConst{\ensuremath{\mathbf{0}}} \HOLSymConst{\ensuremath{\mid}} \HOLConst{\ensuremath{\mathbf{0}}}
\end{alltt}
